Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-15 paper/ms.tex
some notes about implementation

\usepackage{url}
\usepackage{amssymb,amsmath}
\usepackage{natbib}
\usepackage{multirow}
\bibliographystyle{aasjournal}

@@

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
function is
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse will be dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number of
components in the mixture and $N$ is still the number of data points.

@@

>>>>

\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{multirow}
\bibliographystyle{aasjournal}

@@

% Matrix fix:
% http://tex.stackexchange.com/questions/317824/letter-c-appearing-inside-pmatrix-environment-with-aastex
\makeatletter
\def\env@matrix{\hskip -\arraycolsep % taken from amsmath.sty lines 895ff
  \let\@ifnextchar\new@ifnextchar
  \array{*{\c@MaxMatrixCols}c}}
\makeatother

% Column spacing in matrix
% http://tex.stackexchange.com/questions/275725/adjusting-separation-between-matrix-entries
\setlength\arraycolsep{25pt}

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
function is
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse will be dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number of
components in the mixture and $N$ is still the number of data points.

@@

\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
elements are given by
\begin{eqnarray}
K_{nm} = k(\tau_{nm})
\end{eqnarray}
for $k(\tau_{nm})$ given by equation \eq{kernel-simple} could be computed
efficiently by taking advantage of the structure of this covariance function
and \citet{Ambikasaran:2015} generalized this computation to apply to the full
mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
efficient method for computing the determinant of $K$.

\subsection{The algorithm}

To provide some insight for this method, we will start by working through the
simplest possible example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
uncertainties $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate motivate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$.
In matrix notation, this can be written as
\begin{eqnarray}
K\,\bvec{x} &=& \bvec{y} \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ x_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
Now, if we introduce the variables
\begin{eqnarray}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}\eqlabel{algo-first}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,x_1 + e^{c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{c\,t_2}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{c\,t_1} & 0 & 0 & 0 & 0 & 0 \\
    e^{c\,t_1} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & e^{-c\,t_2} & a & 0 & 0 \\
    0 & 0 & e^{-c\,t_2} & a+{\sigma_2}^2 & e^{c\,t_2} & 0 & 0 \\
    0 & 0 & a & e^{c\,t_2} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & e^{-c\,t_3} \\
    0 & 0 & 0 & 0 & 0 & e^{-c\,t_3} & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
